---
output:
  html_document:
    keep_md: yes
  word_document: default
---
# Assessing the Damage
### An analysis of the effects of extreme weather by measuring human and econmic damage
##### Jim Leach 
*******************************************************************************

### Synopsis
This report has been created for an assigment set as part of the *Reproducible Research* MOOC from Johns Hopkins university on [Coursera](https://class.coursera.org/repdata-004). 

The goal was to take historical weather event data and perform some brief exploratory analysis using R, addressing:

* Across the United States, which types of events are most harmful with respect to population health?
* Across the United States, which types of events have the greatest economic consequences?

An additional overall comparison combining human and economic consequences has been performed, using data from Wikipedia on the [value of life](http://en.wikipedia.org/wiki/Value_of_life). In each case, only the results for the top 10 events with the most impact have been presented.

This report will address each of these points and is structured as follows:

* A brief discussion of the analysis and a justification for the chosen methods.
* An overview of the analysis methods.
* A presentation of the results.

********************************************************************************

### Prefacing Discussion of Analysis

It is the case that this analysis is not as full or as accurate as it would be if real insights were required from the data provided.

The measuring of which events caused the most damage on an absolute scale (via total damage caused) was chosen merely for simplicity. A "real" analysis would need to consider many other factors: occurence rate, geographical distribution and more accurate modelling of what constitutes a human "cost" to name but a few.

The analysis has been conducted simply but was performed in a manner that was required for the task at hand - to conduct the analysis in a *reproducible* manner. It was felt that addressing the questions simply and demonstrating that the principles of reproducibility layed out in the course content was more important than a thorough and detailed analysis.


#### Reproducible Analysis

There are a number of factors that demonstrate the *reproducibility* of this report:

* Package installation and loading is conditional - packages are installed and loaded as required.
* The code contains steps to download and read in data. This limits manual steps.
* The code is OS-neutral: wherever possible, it uses conditional logic to ensure compatibility with a range of operating systems.
* Results are generated and stored to variables or external files automatically, so can be recalled later.
* This report has written using RMarkdown and will automatically generate figures and report results - nothing has been hard-coded.


*******************************************************************************
### Data Processing

The methodology of the analysis was to break the code up in to discrete steps. These steps were:

0. Prepare the R working environment
1. Obtain and load the data;
2. Process the data;
3. Summarise events by human cost;
4. Summarise events by economic cost; and
5. Summarise events by human and economic cost combined.

The following sections set out the detailed approach for each stage of the analysis.

##### Step 0 - Preparing the R working environment and package loading

A number of packages were used and are therefore required for this analysis. These packages were:

* [dplyr]((http://blog.rstudio.org/2014/01/17/introducing-dplyr/))
* [ggplot2](http://ggplot2.org/)
* [R.Utils](http://cran.r-project.org/web/packages/R.utils/R.utils.pdf)
* [reshape2](http://seananderson.ca/2013/10/19/reshape.html)

Conditional installation and subsequent loading of these packages was achieved with:

    if(require("package name here",quietly=T)){
      print("loading "package name here"")
    } else {
      print("trying to install "package name here"")
      install.packages(""package name here"")
      library("package name here",quietly=T)
      print(""package name here" installed and loaded")
      if(require("package name here",quietly=T)){
        } else {
        stop("could not install "package name here"")
      }
    }
    
For more information about any of these packages, see their documentation using `help(package="package name here")`.
        
```{r load_paclages, echo=FALSE,results='hide',message=FALSE}
library(dplyr,quietly=T)
library(ggplot2,quietly=T)
library(R.utils,quietly=T)
library(reshape2,quietly=T) 
```

Next, a directory to store the data is created in the working directory:
```{r create_data_dir}
if(!file.exists("./data")){dir.create("./data",showWarnings = FALSE)}
```

##### Step 1i - Getting and loading the data

The data for this analysis are stored [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) and more information about them is available on National Weather Service [page](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf). The download and loading of the data in to R was automated:

```{r get_data, results='hide', warning=FALSE, cache=TRUE}
# 1a. Set the URL for the data
url<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

# 1b. Download and unzip the data - conditional on operating system
if(Sys.info()["sysname"]=="Windows"){setInternet2(use = TRUE)}
if(!file.exists("./data/repdata_data_StormData.csv.bz2"))
{
  ifelse(Sys.info()["sysname"]=="Windows",
         download.file(url,"./data/repdata_data_StormData.csv.bz2",quiet=TRUE),
         download.file(url,"./data/repdata_data_StormData.csv.bz2",method="curl"
                       ,quiet=TRUE)
  )
  bunzip2("./data/repdata_data_StormData.csv.bz2",
          "./data/repdata-data-StormData.csv", overwrite = TRUE, remove = FALSE)
}
# 1c. Read in the data
data<-read.csv("./data/repdata-data-StormData.csv",
               header=T,
               stringsAsFactors=F
              )
```

##### Step 1ii - Initial exploration of the data

Initially, the use of the `str()` function was used to get a sense of the data:
```{r str_data}
str(data)
```

The assigment instructions state that the `evtype` field provides the type of weather event for each record. 

Further, considering both the data dictionary [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) and the questions to be addressed, the following fields in the data were required:

* `fatalities` - the number of fatalities from in the event
* `injuries` - the number of injuries from each event
* `propdmg` - the dollar value of property damage in each event
* `propdmgexp` - the exponent of the property damage field
* `cropdmg` - the dollar value of crop damage in each event
* `crpodmgexp` - the exponent of the crop damage field

###### Exploring `evtype`

An overview of the top 25 event types ranked by number of records showed that there are significant issues with clarity in the data. For example, there appeared to be several versions of the "thunderstorm wind" event type:

```{r}
data %.%
group_by(EVTYPE) %.%
summarise(count = n(), prop = 100*count/nrow(data)) %.%
arrange(-count) %.%
head(25)
```

As part of this analysis, a cleaning exercise was performed to group several of these like-event types together. This was done in step 2 below.

###### Exploring `propdmgexp`

```{r propdmgexp_summary}
data %.%
  group_by(PROPDMGEXP) %.%
  summarise(count = n()) %.%
  arrange(-count)
```  

The data directionary stated that the values "k","m" and "b" represent thousands, millions and billions of dollars respctively. Considering the other values, numeric values were assumed to be the exponent, i.e. a "2" would be one hundred, a "3" one thousand etc. All other values were assumed to represent an exponent of 1. Cleaning of these values was performed in step 2.

###### Exploring `cropdmgexp`

```{r cropdmgexp_summary}
data %.%
  group_by(CROPDMGEXP) %.%
  summarise(count = n()) %.%
  arrange(-count)
```

Similar to for `propdmgexp`, the values for `cropdmgexp` were also cleaned in step 2. The same assumptions were made as for `propdmgexp`.


##### Step 2 - Process the data

The data read in to R were cleaned and processed. Initially names and some data fields were set to lowercase, simply to facilitate typing ease:

```{r data_clean_lowcase}
# 2a. make data field names lower case
names(data) <- tolower(names(data))

# 2b. make evtype values all lower
data$evtype <- tolower(data$evtype)
```

Next, various event types were cleaned to a more standard form. This was done to facilitate better grouping of the data later. 

It should be noted that this exercise was not exhaustive and simply cleaned the data to a state sufficient for some of the most common event types:

```{r data_clean_eventtype, cache=TRUE}
# 2c. replace some of the event types to clean up data
data$evtype_clean <- ifelse(data$evtype=="tstm wind","high wind",
                     ifelse(data$evtype=="heat","excessive heat",
                     ifelse(data$evtype=="flood","floods",
                     ifelse(data$evtype=="flash flood","floods",
                     ifelse(data$evtype=="ice storm","winter storm",
                     ifelse(data$evtype=="thunderstorm wind","high wind",
                     ifelse(data$evtype=="hail","winter storm",
                     ifelse(data$evtype=="heavy snow","winter storm",
                     ifelse(data$evtype=="thunderstorm winds","high wind",
                     ifelse(data$evtype=="blizzard","winter storm",
                     ifelse(data$evtype=="wild/forest fire","wildfire",
                     ifelse(data$evtype=="strong wind","high wind",
                     ifelse(data$evtype=="winter storms","winter storm",
                     ifelse(data$evtype=="storm surge","floods",
                     ifelse(data$evtype=="hurricane","hurricane/typhoon",
                     ifelse(data$evtype=="river flood","floods",
                     ifelse(data$evtype=="tropical storm","hurricane/typhoon",
                     ifelse(data$evtype=="heavy rains/severe weather",
                                          "winter storm",
                     ifelse(data$evtype=="heavy rain/severe weather",
                                          "winter storm",
                     ifelse(data$evtype=="storm surge/tide","floods",
                     ifelse(data$evtype=="hurricane opal","hurricane/typhoon",
                     ifelse(data$evtype=="tornadoes, tstm wind, hail",
                                          "winter storm",
                     ifelse(data$evtype=="severe thunderstorm","winter storm",
                     ifelse(data$evtype=="frost/freeze","extreme cold",
                     ifelse(data$evtype=="marine tstm wind","high wind",
                     ifelse(data$evtype=="winter weather","winter storm",
                     ifelse(data$evtype=="marine thunderstorm wind","high wind",
                     data$evtype)))))))))))))))))))))))))))

# 2d. make all uppercase for presentation later
data$evtype_clean <- toupper(data$evtype_clean)
```

The values for the property and crop damage exponents were standardised based on the assumptions set out previously. 

This was done to facilitate more accurate calculations of the economic cost of various events. 

```{r data_clean_exponents, cache=TRUE}
# 2e. recode property damage exponenets
data$propexp_clean <- ifelse(data$propdmgexp=="K",1E3,
                      ifelse(data$propdmgexp=="M",1E6,
                      ifelse(data$propdmgexp=="B",1E9,
                      ifelse(data$propdmgexp=="m",1E6,
                      ifelse(data$propdmgexp=="5",1E5,
                      ifelse(data$propdmgexp=="2",1E2,
                      ifelse(data$propdmgexp=="7",1E7,
                      ifelse(data$propdmgexp=="3",1E3,
                      ifelse(data$propdmgexp=="4",1E4,
                      ifelse(data$propdmgexp=="6",1E6,
                      ifelse(data$propdmgexp=="8",1E8,1)))))))))))

data$cropexp_clean <- ifelse(data$cropdmgexp=="K",1E3,
                      ifelse(data$cropdmgexp=="k",1E3,
                      ifelse(data$cropdmgexp=="M",1E6,
                      ifelse(data$cropdmgexp=="m",1E6,
                      ifelse(data$cropdmgexp=="B",1e9,
                      ifelse(data$cropdmgexp=="2",1E2,1))))))
```

Finally, a new field was added to the data to list the median value of a human life, as calculated by various US agencies ([source](en.wikipedia.org/wiki/Value_of_life)). This field was used in the combined analysis of human and economic cost.
```{r data_clean_humanvalue, cache=TRUE}
# 2f. esitmation of human value
data$human_cost <- rep(7E6,nrow(data))
```

#####  Step 3 - Summarise by Injury/Death

First, the data were summarised, grouping around the cleaned `evtype` value and summing total deaths and injuries:
```{r ev_sum_summary}
# 3a. summarise by event type
ev_sum <- data %.%
          select(evtype_clean,fatalities,injuries) %.%
          group_by(evtype_clean) %.%
          summarise(deaths=sum(fatalities),
                    injuries=sum(injuries),
                    affected=sum(deaths,injuries)) %.%
          arrange(-affected) %.%
          transform(evtype_clean =
                      reorder(evtype_clean,order(affected,decreasing=T))) %.%
          select(-affected)
```

The resulting data frame was *melted* to a structure suitable for graphing:
```{r ev_sum_melt}
# 3b. melt to assist graphing
ev_sum <- melt(ev_sum[1:10,],
               id.vars="evtype_clean",
               variable.name="efftype",
               value.name="people")

# 3c. set up the factor variable
ev_sum$efftype <- factor(ev_sum$efftype,levels=c("deaths","injuries"))
```

The results of this summary are presented in the results section.

#####  Step 4 -  Summarise by property/crop damage

First, the data were summarised, grouping around the cleaned `evtype` value and calculating the total property and crop damage cost from the provided values and the exponents for each:
```{r ev_sum_eco_summary}
# 4a. summarise by event type
ev_sum_eco <- data %.%
              select(evtype_clean,
                     propdmg,
                     propexp_clean,
                     cropdmg,
                     cropexp_clean) %.%
              group_by(evtype_clean) %.%
              summarise(property=sum(propdmg*propexp_clean), # use exponents
                        crop=sum(cropdmg*cropexp_clean),
                        total_cost=sum(property,crop)) %.%
              arrange(-total_cost) %.%
              transform(evtype_clean =
                          reorder(evtype_clean,order(total_cost,decreasing=T)))%.%
              select(-total_cost)
```

The resulting data frame was *melted* to a structure suitable for graphing:
```{r ev_sum_eco_melt}
# 3b. melt the data frame to suit graphing
ev_sum_eco <- melt(ev_sum_eco[1:10,],
               id.vars="evtype_clean",
               variable.name="dmgtype",
               value.name="cost")
# set up factor levels
ev_sum_eco$dmgtype <- factor(ev_sum_eco$dmgtype,levels=c("crop","property"))
# create a field in millions to aid graphing
ev_sum_eco$cost_clean <- ev_sum_eco$cost/1E6
```

The results of this summary are presented in the results section.

#####  Step 5 -  Combine human and economic cost to rank all events

First, the data were summarised, grouping around the cleaned `evtype` value. The total economic cost was calculate from the provided values and the exponents for each. 

The total human cost was calculated using the value of a human life sourced from Wikipedia. Where a fatality was recorded the full value of a human life was used, for an injury only half of this value was used.

```{r ev_sum_both_summary}
# 5a. summarise by event type
ev_sum_both <- data %.%
               select(evtype_clean,
                      injuries,
                      fatalities,
                      human_cost,
                      propdmg,
                      propexp_clean,
                      cropdmg,
                      cropexp_clean) %.%
                group_by(evtype_clean) %.%
                summarise(
                human_cost = sum((fatalities*human_cost)+(0.5*(injuries*human_cost))),
                eco_cost = sum((propdmg*propexp_clean)+(cropdmg*cropexp_clean)),
                total_cost = sum(human_cost,eco_cost)
                ) %.%
                arrange(-total_cost) %.%
                transform(evtype_clean =
                            reorder(evtype_clean,order(total_cost,decreasing=T))) %.%
                select(-total_cost)
```

The resulting data frame was *melted* to a structure suitable for graphing:
```{r ev_sum_both_melt}
# 5b. melt the data frame to suit graphing
ev_sum_both <- melt(ev_sum_both[1:10,],
                   id.vars="evtype_clean",
                   variable.name="efftype",
                   value.name="cost")
# set up factor levels
ev_sum_both$efftype <- factor(ev_sum_both$efftype,levels=c("human_cost","eco_cost"))
# create a field in millions to aid graphing
ev_sum_both$cost_clean <- ev_sum_both$cost/1E6
```

The results of this summary are presented in the results section.

*******************************************************************************
### Results
#### Summary of Human Cost
##### Figure 1 - Top 10 Results ranked by total individuals affected
```{r human_cost_plot}
# create summary for text underneath figure
human <- ev_sum %.%
         group_by(evtype_clean) %.%
         summarise(total=sum(people)) %.%
         arrange(-total)
human_ev <- as.character(human[1,1])
human_ev_count <- human[1,2]
rm(human)

# create figure
options(scipen=100)
ggplot(data=ev_sum,aes(x=evtype_clean,y=people,fill=efftype))+
geom_bar(stat="identity")+
theme_minimal()+
scale_fill_manual(values=c("firebrick2","firebrick"), name = element_blank(),
                 breaks = c("deaths","injuries"),labels=c("Deaths","Injuries"))+
ylab("Individual")+
ggtitle("Total human cost per event type")+
scale_y_continuous(breaks=seq(from=0,
                              to=(max(ev_sum$people)+0.1*max(ev_sum$people)),
                              by=10000))+
theme(axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major.x = element_blank(),
      legend.position="bottom",
      axis.title.x=element_blank())
```

The above figure shows that `r I(human_ev)` is the most harmful event with respect to human injuries a deaths. In total, over the entire data collection period `r I(human_ev_count)` individuals were injured or killed.


#### Summary of Economic Cost
##### Figure 2 - Top 10 Results ranked by total cost of property and crop damage
```{r eco_cost_plot}
# create summary for text underneath figure
eco <- ev_sum_eco %.%
       group_by(evtype_clean) %.%
       summarise(total=sum(cost_clean)) %.%
       arrange(-total)
eco_ev <- as.character(eco[1,1])
eco_ev_count <- round(eco[1,2])
rm(eco)

# create figure
options(scipen=100)
ggplot(data=ev_sum_eco,aes(x=evtype_clean,y=cost_clean,fill=dmgtype))+
geom_bar(stat="identity")+
theme_minimal()+
scale_fill_manual(values=c("goldenrod","gold"),name = element_blank(),
                  breaks = c("crop","property"),
                  labels=c("Crop Damage","Property Damage"))+
ylab("m$")+
ggtitle("Total property cost per event type")+
scale_y_continuous(breaks=seq(from=0,
                              to=(max(ev_sum_eco$cost)+0.1*max(ev_sum_eco$cost)),
                              by=25000))+
theme(axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major.x=element_blank(),
      legend.position="bottom",
      axis.title.x=element_blank())
```

The above figure shows that `r I(eco_ev)` is the most harmful event with respect to property and crop damage. In total, over the entire data collection period `r I(eco_ev_count)` m$ worth of damage was caused.

#### Summary of Combined Cost
##### Figure 3 - Top 10 Results ranked by total combined economic and human cost
```{r both_cost_plot}
# create summary for text underneath figure
eco <- ev_sum_both %.%
       group_by(evtype_clean) %.%
       summarise(total=sum(cost_clean)) %.%
       arrange(-total)
both_ev <- as.character(eco[1,1])
both_ev_count <- round(eco[1,2])
rm(eco)

# create figure
options(scipen=100)
ggplot(data=ev_sum_both,aes(x=evtype_clean,y=cost_clean,fill=efftype))+
geom_bar(stat="identity")+
theme_minimal()+
scale_fill_manual(values=c("firebrick","goldenrod"),
                  name = element_blank(),
                  breaks = c("human_cost","eco_cost"),
                  labels=c("Human Cost","Property Cost"))+
ylab("m$")+
ggtitle("Total combined human and property cost per event type")+
scale_y_continuous(breaks=seq(
  from=0,to=(max(ev_sum_both$cost)+0.1*max(ev_sum_both$cost)),by=25000)
)+
theme(axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major.x=element_blank(),
      legend.position="bottom",
      axis.title.x=element_blank()
)
```

The above figure shows that `r I(both_ev)` is the most harmful event with respect human and economic costs. In total, over the entire data collection period `r I(both_ev_count)` m$ worth of damage was caused to humans and property.

#### Overall Conclusion

It is the case that measured by human cost alone, the `r I(human_ev)` is the most dangerous event with respect to human health. `r I(eco_ev)` have caused the most damage to crops and property. Combining an approximate dollar value for human life with economic dollar cost, it is the case that `r I(both_ev)` is ranked as the most dangerous. 

For the purposes of this simplistic analysis it is clear that, on a combined scale, `r I(both_ev)` is the most significant event. However, this analysis has not controlled for a very large number of factors and further work and analysis would be required to obtain a more accurate and meaningful measure of event severity.


********************************************************************************

### References and Contact

The complete code and documentation for this assignment can be found on [GitHub](https://github.com/Jim89/reproducible2) repository. 

The author of this report can be contacted on [twitter](https://twitter.com/leach_jim) 

********************************************************************************
